---
title: "Typha_Fungi_Taxa_Analysis"
author: "Saysha Sebren and Charlie Bond"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---
# Typha Microbial Community Analysis
### Introduction
In the previous R markdown for ITS bioinformatics (DADA2 pipeline for processing amplicon sequence data), we generated ASV counts and taxonomy for fungi (ITS). Now we will take our first look at the data, the number of ASVs and taxonomic groups, and the occurrences of ASV in each location site over 3 time points. 

## Setup and set working directory
```{r, include=FALSE}
rm(list = ls())
knitr::opts_chunk$set(echo = TRUE)
# Set working directory
setwd("C:/Users/saysh/OneDrive/Desktop/Kuehn_Final_Analyses/Typha/Typha_Fungi/Typha_Fungi_Rerun_March2026")
knitr::opts_knit$set(root.dir = "C:/Users/saysh/OneDrive/Desktop/Kuehn_Final_Analyses/Typha/Typha_Fungi/Typha_Fungi_Rerun_March2026")
options(knitr.duplicate.label = "allow") ### so I don't have to give chunks unique names
```

## Load libraries
```{r, library}
# Load necessary libraries
library(phyloseq)
library(vegan)
library(ggplot2)
library(tidyr)
library(dplyr)
library(pheatmap)
library(GUniFrac)
library(raster)
library(pals)
library(RColorBrewer)
library(ragg)
library(ggpubr)
```

## Set Theme
```{r, theme}
# Set theme
theme_set(theme_bw() + theme(
              plot.title = element_text(size=15, color="black"),
              axis.text.x = element_text(size=15, color="black"),
              axis.text.y = element_text(size=15, color="black"),
              axis.title.x = element_text(size=15),
              axis.title.y = element_text(size=15),
              legend.text = element_text(size=12),
              legend.title = element_text(size=15),
              panel.background = element_blank(), 
              panel.border = element_rect(colour = "black", fill=NA, size=1),
              plot.background = element_blank()))
```

# Statistical Analysis
## Phyloseq
```{r}
# Set working directory
setwd("C:/Users/saysh/OneDrive/Desktop/Kuehn_Final_Analyses/Typha/Typha_Fungi/Typha_Fungi_Rerun_March2026")

# Create phyloseq object
# Read in ASV table
asvtest <- read.table("dada2_Typha_ITS_asvtable_061526.count_table", header=T, row.names=1, check.names=F, sep="\t")
rownames(asvtest)

# Read in taxonomy table
taxtest <- read.csv("Typha_final_fungi_taxtab_061926.csv", row.names=1)
rownames(taxtest)

# Read in metadata table
metatest <- read.csv("KevinTypha_metadata.csv", row.names=1)
rownames(metatest)

# Separate metadata by location (A = Weeks Bay, Alabama; M = Michigan)
metatestA <- metatest %>% filter(Location =="Alabama")
metatestA
metatestM <- metatest %>% filter(Location =="Michigan")
metatestM

# Designate file types
asvtest <- otu_table(asvtest, taxa_are_rows = FALSE)
taxtest <- tax_table(as.matrix(taxtest))
metatest <- sample_data(metatest)

# Separate by location
metatestA <- sample_data(metatestA)
metatestM <- sample_data(metatestM)

# Combine into phyloseq object
pseqtest <- phyloseq(asvtest, taxtest, metatest)
pseqtestA <- phyloseq(asvtest, taxtest, metatestA)
pseqtestM <- phyloseq(asvtest, taxtest, metatestM)

# Extract easy-to-use sample data
samdftest <- data.frame(sample_data(pseqtest))
samdftestA <- data.frame(sample_data(pseqtestA))
samdftestM <- data.frame(sample_data(pseqtestM))
```

```{r}
## Remove singletons, doubletons, and ASVs with less than 5 reads across all samples. 
cured_asvsITS<- pseqtest@otu_table
summary(rowSums(cured_asvsITS))
sum(rowSums(cured_asvsITS))
ncol(cured_asvsITS)

cured_asvsITS <- cured_asvsITS[,colSums(cured_asvsITS)>5]
summary(rowSums(cured_asvsITS))
rowSums(cured_asvsITS)

# Update phyloseq objects
pseqtest <- phyloseq(cured_asvsITS, taxtest, metatest)
pseqtestA <- subset_samples(pseqtest, Location == "Alabama")
pseqtestM <- subset_samples(pseqtest, Location == "Michigan")
```


## Remove taxa indicators on names to streamline analysis
```{r}
setwd("C:/Users/saysh/OneDrive/Desktop/Kuehn_Final_Analyses/Typha/Typha_Fungi/Typha_Fungi_Rerun_March2026")
library(stringr)
tax_table(pseqtest)[,"Species"] <- sapply(str_replace(tax_table(pseqtest)[,"Species"], "s__", ""),`[`, 1)
tax_table(pseqtest)[,"Genus"] <- sapply(str_replace(tax_table(pseqtest)[,"Genus"], "g__", ""),`[`, 1)
tax_table(pseqtest)[,"Family"] <- sapply(str_replace(tax_table(pseqtest)[,"Family"], "f__", ""),`[`, 1)
tax_table(pseqtest)[,"Order"] <- sapply(str_replace(tax_table(pseqtest)[,"Order"], "o__", ""),`[`, 1)
tax_table(pseqtest)[,"Class"] <- sapply(str_replace(tax_table(pseqtest)[,"Class"], "c__", ""),`[`, 1)
tax_table(pseqtest)[,"Phylum"] <- sapply(str_replace(tax_table(pseqtest)[,"Phylum"], "p__", ""),`[`, 1)
tax_table(pseqtest)[,"Kingdom"] <- sapply(str_replace(tax_table(pseqtest)[,"Kingdom"], "k__", ""),`[`, 1)
```

## Preliminary taxonomic overview:
```{r}
setwd("C:/Users/saysh/OneDrive/Desktop/Kuehn_Final_Analyses/Typha/Typha_Fungi/Typha_Fungi_Rerun_March2026")
### Extract taxonomy table from phyloseq object
taxtabITS<- as.data.frame(tax_table(pseqtest))

## Number of fungal ASVs
nrow(taxtabITS[taxtabITS$Kingdom=="Fungi",])

## Unique phyla of Fungi
x<- unique(taxtabITS[taxtabITS$Kingdom=="Fungi" & !is.na(taxtabITS[,2]),2])
length(x)
x

## Unique classes of Fungi
x<- unique(taxtabITS[taxtabITS$Kingdom=="Fungi" & !is.na(taxtabITS[,3]),3])
length(x)
x

## Unique orders of Fungi
x<- unique(taxtabITS[taxtabITS$Kingdom=="Fungi" & !is.na(taxtabITS[,4]),4])
length(x)
x

## Unique families of Fungi
x<- unique(taxtabITS[taxtabITS$Kingdom=="Fungi" & !is.na(taxtabITS[,5]),5])
length(x)

## Unique genera of Fungi
x<- unique(taxtabITS[taxtabITS$Kingdom=="Fungi" & !is.na(taxtabITS[,6]),6])
length(x)

## Number of Fungal ASVs without family assigned
sum(is.na(taxtabITS[taxtabITS$Kingdom=="Fungi",5]))

## Number of Fungal ASVs without genus assigned
sum(is.na(taxtabITS[taxtabITS$Kingdom=="Fungi",6]))

## Proportion of Fungal ASVs without genus assigned
sum(is.na(taxtabITS[taxtabITS$Kingdom=="Fungi",6]))/nrow(taxtabITS[taxtabITS$Kingdom=="Fungi",])

```

## Beta Diversity
```{r}
setwd("C:/Users/saysh/OneDrive/Desktop/Kuehn_Final_Analyses/Typha/Typha_Fungi/Typha_Fungi_Rerun_March2026")
# Plot species accumulation curve
Aotu <- pseqtest@otu_table[samdftest$Location=='Alabama',]
Motu <- pseqtest@otu_table[samdftest$Location=='Michigan',]

spaccumA <- specaccum(Aotu, method ="random")
spaccumM <- specaccum(Motu, method ="random")
plot(spaccumA)
plot(spaccumM)

# Generate distance matrices
braytest <- phyloseq::distance(pseqtest, method = "bray")
braytestA <- phyloseq::distance(pseqtestA, method = "bray")
braytestM <- phyloseq::distance(pseqtestM, method = "bray")

# Bray-Curtis distance matrices for NMDS graphs
braymdsA <- metaMDS(braytestA, k=2, trymax=500, wascores = T)
braymdsM <- metaMDS(braytestM, k=2, trymax=500, wascores = T)
brayscoresA <- scores(braymdsA)
brayscoresM <- scores(braymdsM)
braymdsA
braymdsM
```

## Permutational Multivariate Analysis of Variance (PERMANOVA)

```{r}
setwd("C:/Users/saysh/OneDrive/Desktop/Kuehn_Final_Analyses/Typha/Typha_Fungi/Typha_Fungi_Rerun_March2026")
# PERMANOVA
names(samdftest)
adonis2(braytest ~ Location*Date, data = samdftest)
```

# Analysis of Similarity (ANOSIM)
## ANOSIM for both sites

```{r}
setwd("C:/Users/saysh/OneDrive/Desktop/Kuehn_Final_Analyses/Typha/Typha_Fungi/Typha_Fungi_Rerun_March2026")
# Combine metadata and ASVs for ANOSIM
KevinTyphaFungi_full_data <- cbind(metatest,asvtest)

# Calculate proportional data
asv_prop <- asvtest/apply(asvtest,1,sum)
index <- seq(1, 36, by=2)

# ANOSIM testing for differences among seasons using a Bray-Curtis distance matrix
dist <- vegdist(asv_prop, method='bray')
Date_ano <- anosim(dist, KevinTyphaFungi_full_data$Date)
summary(Date_ano)
```

## ANOSIM for Alabama

```{r}
setwd("C:/Users/saysh/OneDrive/Desktop/Kuehn_Final_Analyses/Typha/Typha_Fungi/Typha_Fungi_Rerun_March2026")
# Select for Alabama samples
KevinTyphaFungi_full_dataA <- KevinTyphaFungi_full_data[seq(2,36,by=2),]

# Calculate proportional data
indexA <- seq(2, 36, by=2)
asv_propA <- asv_prop[seq(2,36,by=2),]

# Double check for correct sample rows
sum(rownames(asv_propA) == rownames(KevinTyphaFungi_full_dataA))

# ANOSIM testing for differences among seasons using a Bray-Curtis distance matrix
distA <- vegdist(asv_propA, method='bray')
Date_anoA <- anosim(distA, KevinTyphaFungi_full_dataA$Date)
summary(Date_anoA)
```

## ANOSIM for Michigan

```{r}
setwd("C:/Users/saysh/OneDrive/Desktop/Kuehn_Final_Analyses/Typha/Typha_Fungi/Typha_Fungi_Rerun_March2026")
# Select for Michigan samples
KevinTyphaFungi_full_dataM <- KevinTyphaFungi_full_data[seq(1,36,by=2),]

# Calculate proportional data
indexM <- seq(1, 36, by=2)
asv_propM <- asv_prop[seq(1,36,by=2),]

# Double check for correct sample rows
sum(rownames(asv_propM) == rownames(KevinTyphaFungi_full_dataM))

# ANOSIM testing for differences among seasons using a Bray-Curtis distance matrix
distM <- vegdist(asv_propM, method='bray')
Date_anoM <- anosim(distM, KevinTyphaFungi_full_dataM$Date)
summary(Date_anoM)
```

## Alpha Diversity and MANOVA
```{r}
setwd("C:/Users/saysh/OneDrive/Desktop/Kuehn_Final_Analyses/Typha/Typha_Fungi/Typha_Fungi_Rerun_March2026")

# Calculate alpha diversity
alpha <- estimate_richness(pseqtest, split=TRUE, measures=NULL)
row.names(alpha) <- row.names(samdftest)

# Save alpha diversity
write.csv(alpha, "KuehnTyphaFungi_alpha_diversity_062626.csv")

# Merge alpha diversity and metadata
alpha <- merge(metatest, alpha, by="row.names", all=TRUE)
alpha

# Multivariate analysis of variance (MANOVA)
man1 <- manova(cbind(Observed, Chao1, Shannon, InvSimpson) ~ Location*Date, data=alpha)
hist(man1$residuals)
plot(man1$residuals ~ man1$fitted.values)
summary.aov(man1)
```
# Taxonomy
## Phylum count table
```{r}
setwd("C:/Users/saysh/OneDrive/Desktop/Kuehn_Final_Analyses/Typha/Typha_Fungi/Typha_Fungi_Rerun_March2026")
# Make taxonomy object by phylum
phyla_counts_tab <- otu_table(tax_glom(pseqtest, taxrank="Phylum"), taxa_are_rows = FALSE)
phyla_counts_tab <- t(phyla_counts_tab)

# Make vector of phyla names to set as row names
phyla_tax_vec <- as.vector(tax_table(tax_glom(pseqtest, taxrank="Phylum"))[,2]) 
rownames(phyla_counts_tab) <- as.vector(phyla_tax_vec)
phyla_counts_tab
write.csv(phyla_counts_tab, "phyla_counts_062626.csv")

asv_counts <-pseqtest@otu_table
# Determine the number of unclassified sequences at the phylum level
Unclassified_Phylum <- colSums(t(asv_counts)) - colSums(phyla_counts_tab)

# Add a row of "unclassified" to the phylum count table
phyla_and_unidentified_counts_tab <- rbind(phyla_counts_tab, 
"Unclassified_Phylum"=Unclassified_Phylum)
phyla_and_unidentified_counts_tab

# Check to see if the phylum and ASV tables have identical data
identical(colSums(phyla_and_unidentified_counts_tab), rowSums(asv_counts))
```
## Class count table
```{r}
setwd("C:/Users/saysh/OneDrive/Desktop/Kuehn_Final_Analyses/Typha/Typha_Fungi/Typha_Fungi_Rerun_March2026")
# Make count table broken down by class
class_counts_tab <- otu_table(tax_glom(t(pseqtest), taxrank="Class"))

# Make table containing phylum and class data
class_tax_phy_tab <- tax_table(tax_glom(t(pseqtest), taxrank="Class"))
phy_tmp_vec <- class_tax_phy_tab[,2]
class_tmp_vec <- class_tax_phy_tab[,3]
rows_tmp <- row.names(class_tax_phy_tab)
class_tax_tab <- data.frame("Phylum"=phy_tmp_vec, "Class"=class_tmp_vec, 
row.names = rows_tmp)
class_tax_tab
```

## Make separate tables
```{r}
# Remove Ascomycota to separate it into classes later
basidio_temp_major_taxa_counts_tab <- phyla_counts_tab[!row.names(phyla_counts_tab) %in% "Ascomycota",]
# Remove Basidiomycota to separate into classes later
asco_temp_major_taxa_counts_tab <- phyla_counts_tab[!row.names(phyla_counts_tab) %in% "Basidiomycota",]
```


## Raw Abundance
```{r}
setwd("C:/Users/saysh/OneDrive/Desktop/Kuehn_Final_Analyses/Typha/Typha_Fungi/Typha_Fungi_Rerun_March2026")
# Make vector of just the Ascomycota classes
asco_classes_vec <- as.vector(class_tax_tab[class_tax_tab$Phylum == "Ascomycota", "Class"])
row.names(class_counts_tab) <- as.vector(class_tax_tab$Class)

# Make table of Ascomycota classes
asco_class_counts_tab <- class_counts_tab[row.names(class_counts_tab) %in% asco_classes_vec,]

# Make table of Ascomycota not identified to the class level
asco_no_class_annotated_counts <- phyla_and_unidentified_counts_tab[row.names(phyla_and_unidentified_counts_tab) %in% "Ascomycota",]- colSums(asco_class_counts_tab)
```

```{r}
# Make vector of just the Basidiomycota classes
basidio_classes_vec <- as.vector(class_tax_tab[class_tax_tab$Phylum == "Basidiomycota", "Class"])
row.names(class_counts_tab) <- as.vector(class_tax_tab$Class)

# Make table of Basidiomycota classes
basidio_class_counts_tab <- class_counts_tab[row.names(class_counts_tab) %in% basidio_classes_vec,]

# Make table of Basidiomycota not identified to the class level
basidio_no_class_annotated_counts <- phyla_and_unidentified_counts_tab[row.names(phyla_and_unidentified_counts_tab) %in% "Basidiomycota",]- colSums(basidio_class_counts_tab)
```

```{r}
# Now combine the tables and the ASVs with an Unclassified Phylum
major_taxa_counts_tab <- rbind(Unclassified_Phylum, asco_class_counts_tab,
 "Unclassified_Ascomycota"=asco_no_class_annotated_counts, basidio_class_counts_tab, "Unclassified_Basidiomycota"=basidio_no_class_annotated_counts)
major_taxa_counts_tab

# Save
write.csv(major_taxa_counts_tab, "KevinTyphaFungi_PhylumRawAbundance_062626.csv")

# Combining all tables to see both Phyla and Classes
major_taxa_counts_tab_full <- rbind(asco_temp_major_taxa_counts_tab, basidio_temp_major_taxa_counts_tab,
                               Unclassified_Phylum, asco_class_counts_tab,
 "Unclassified_Ascomycota"=asco_no_class_annotated_counts, basidio_class_counts_tab, "Unclassified_Basidiomycota"=basidio_no_class_annotated_counts)
major_taxa_counts_tab_full

# Save
write.csv("Typha_MajorTaxaCounts_PhylaClass_062626.csv")
```


## Relative Abundance

```{r}
setwd("C:/Users/saysh/OneDrive/Desktop/Kuehn_Final_Analyses/Typha/Typha_Fungi/Typha_Fungi_Rerun_March2026")
# Check that all sequences are accounted for
identical(colSums(major_taxa_counts_tab), rowSums(asv_counts))

# Convert totals to relative abundance
major_taxa_proportions_tab <- apply(major_taxa_counts_tab, 2, function(x) 
x/sum(x)*100)

# Save Relative Abundance
write.csv(major_taxa_proportions_tab, "Typha_ITS_RelativeAbundance_062626.csv")
```

### Organize Relative Abundance Table
```{r}
# Add totals row
major_taxa_proportions_tab_total <- rbind(major_taxa_proportions_tab, colSums(major_taxa_proportions_tab))

rownames(major_taxa_proportions_tab_total)[nrow(major_taxa_proportions_tab_total)] <- "Total %"

print(major_taxa_proportions_tab_total)

write.csv(major_taxa_proportions_tab_total, "KTF_PhylumRelativeAbundance_total%_062626.csv")

# Divide into two tables based on location
# Alabama - Weeks Bay
PhyRelAbun_A <- major_taxa_proportions_tab[, c("kn1wb", "kn2wb", "kn3wb", "kn4wb", "kn5wb", "kn6wb", "km1wb", "km2wb", "km3wb", "km4wb", "km5wb", "km6wb","kj1wb", "kj2wb", "kj3wb", "kj4wb", "kj5wb", "kj6wb")]
# Michigan
PhyRelAbun_M <- major_taxa_proportions_tab[, c("kn1mi", "kn2mi", "kn3mi", "kn4mi", "kn5mi", "kn6mi", "km1mi", "km2mi", "km3mi", "km4mi", "km5mi", "km6mi","kj1mi", "kj2mi", "kj3mi", "kj4mi", "kj5mi", "kj6mi")]

#Divide each table based on date
# Alabama, November
PhyRelAbun_A_Nov <- major_taxa_proportions_tab[, c("kn1wb", "kn2wb", "kn3wb", "kn4wb", "kn5wb", "kn6wb")]
# Alabama, March
PhyRelAbun_A_Mar <- major_taxa_proportions_tab[, c("km1wb", "km2wb", "km3wb", "km4wb", "km5wb", "km6wb")]
# Alabama, July
PhyRelAbun_A_Jul <- major_taxa_proportions_tab[, c("kj1wb", "kj2wb", "kj3wb", "kj4wb", "kj5wb", "kj6wb")]
# Michigan, November
PhyRelAbun_M_Nov <- major_taxa_proportions_tab[, c("kn1mi", "kn2mi", "kn3mi", "kn4mi", "kn5mi", "kn6mi")]
# Michigan, March
PhyRelAbun_M_Mar <- major_taxa_proportions_tab[, c("km1mi", "km2mi", "km3mi", "km4mi", "km5mi", "km6mi")]
# Michigan, July
PhyRelAbun_M_Jul <- major_taxa_proportions_tab[, c("kj1mi", "kj2mi", "kj3mi", "kj4mi", "kj5mi", "kj6mi")]

library(matrixStats)
# Get the Mean and Standard Deviation of each row
# Alabama, November
MeanAN <- rowMeans(PhyRelAbun_A_Nov, na.rm = TRUE)
MeanAN <- cbind(PhyRelAbun_A_Nov, MeanAN) 
StDevAN <- apply(PhyRelAbun_A_Nov, 1, sd)
RelAbun_A_Nov_MeanSD <- cbind(MeanAN, StDevAN)
# Alabama, March
MeanAM <- rowMeans(PhyRelAbun_A_Mar, na.rm = TRUE)
MeanAM <- cbind(PhyRelAbun_A_Mar, MeanAM) 
StDevAM <- apply(PhyRelAbun_A_Mar, 1, sd)
RelAbun_A_Mar_MeanSD <- cbind(MeanAM, StDevAM)
# Alabama, July
MeanAJ <- rowMeans(PhyRelAbun_A_Jul, na.rm = TRUE)
MeanAJ <- cbind(PhyRelAbun_A_Jul, MeanAJ) 
StDevAJ <- apply(PhyRelAbun_A_Jul, 1, sd)
RelAbun_A_Jul_MeanSD <- cbind(MeanAJ, StDevAJ)
# Michigan, November
MeanMN <- rowMeans(PhyRelAbun_M_Nov, na.rm = TRUE)
MeanMN <- cbind(PhyRelAbun_M_Nov, MeanMN) 
StDevMN <- apply(PhyRelAbun_M_Nov, 1, sd)
RelAbun_M_Nov_MeanSD <- cbind(MeanMN, StDevMN)
# Michigan, March
MeanMM <- rowMeans(PhyRelAbun_M_Mar, na.rm = TRUE)
MeanMM <- cbind(PhyRelAbun_M_Mar, MeanMM) 
StDevMM <- apply(PhyRelAbun_M_Mar, 1, sd)
RelAbun_M_Mar_MeanSD <- cbind(MeanMM, StDevMM)
# Michigan, July
MeanMJ <- rowMeans(PhyRelAbun_M_Jul, na.rm = TRUE)
MeanMJ <- cbind(PhyRelAbun_M_Jul, MeanMJ) 
StDevMJ <- apply(PhyRelAbun_M_Jul, 1, sd)
RelAbun_M_Jul_MeanSD <- cbind(MeanMJ, StDevMJ)
# Save Relative Abundance Tables
write.csv(RelAbun_A_Nov_MeanSD, "RelAbun_A_Nov_MeanSD_062626.csv")
write.csv(RelAbun_A_Mar_MeanSD, "RelAbun_A_Mar_MeanSD_062626.csv")
write.csv(RelAbun_A_Jul_MeanSD, "RelAbun_A_Jul_MeanSD_062626.csv")
write.csv(RelAbun_M_Nov_MeanSD, "RelAbun_M_Nov_MeanSD_062626.csv")
write.csv(RelAbun_M_Mar_MeanSD, "RelAbun_M_Mar_MeanSD_062626.csv")
write.csv(RelAbun_M_Jul_MeanSD, "RelAbun_M_Jul_MeanSD_062626.csv")
```


### Figure Generation

## NMDS Ordination

```{r}
#Date <- as.factor(KevinTyphaFungi_full_data$Date)
setwd("C:/Users/saysh/OneDrive/Desktop/Kuehn_Final_Analyses/Typha/Typha_Fungi/Typha_Fungi_Rerun_March2026")

# Plot Alabama NMDS
pdf("KevinTyphaFungi_NMDSplotA_062626.pdf")
nmdsp1A <- plot_ordination(pseqtestA, braymdsA, type="samples", color="Date")+ 
  geom_point(size=2)+ stat_ellipse()+
  geom_hline(yintercept = 0, color = "black", linetype = "solid", linewidth = 0.5) + 
  # Add vertical line at x=0
  geom_vline(xintercept = 0, color = "black", linetype = "solid", linewidth = 0.5) +
  theme_minimal() # Or your preferred theme
dev.off()
nmdsp1A
nmdsp1A_data <- nmdsp1A$data
print(nmdsp1A_data)
write.csv(nmdsp1A_data, "NMDScoord_KuehnTyphaFungi_Alabama_062626.csv")

# Alternate way to plot
plotout_nmdsp1A <- "KevinTyphaFungi_NMDSplotA_062626.tiff"
agg_tiff(filename=plotout_nmdsp1A, width=3200, height=3000, units="px",
         pointsize=10, res=600, compression="lzw", scaling=0.5)
nmdsp1A
invisible(dev.off())

# Plot Michigan NMDS
pdf("KevinTyphaFungi_NMDSplotM_062626.pdf")
nmdsp1M <- plot_ordination(pseqtestM, braymdsM, type="samples", color="Date")+ 
  geom_point(size=2)+ stat_ellipse()+
  geom_hline(yintercept = 0, color = "black", linetype = "solid", linewidth = 0.5) + 
  # Add vertical line at x=0
  geom_vline(xintercept = 0, color = "black", linetype = "solid", linewidth = 0.5) +
  theme_minimal() # Or your preferred theme
dev.off()
nmdsp1M
nmdsp1M_data <- nmdsp1M$data
print(nmdsp1M_data)
write.csv(nmdsp1M_data, "NMDScoord_KuehnTyphaFungi_Michigan_062626.csv")

# Alternate way to plot
plotout_nmdsp1M <- "KevinTyphaFungi_NMDSplotM_062626.tiff"
agg_tiff(filename=plotout_nmdsp1M, width=3200, height=3000, units="px",
         pointsize=10, res=600, compression="lzw", scaling=0.5)
nmdsp1M
invisible(dev.off())
```

## Network Analysis

```{r}
setwd("C:/Users/saysh/OneDrive/Desktop/Kuehn_Final_Analyses/Typha/Typha_Fungi/Typha_Fungi_Rerun_March2026")
#Network analysis
net1A <- make_network(pseqtestA, type="samples", distance="bray", max.dist=0.7, 
keep.isolates=TRUE)
netp1A <- plot_network(net1A, pseqtestA, type="samples", color="Date", 
#shape="Location",
 point_size=3, label=NA)+
 theme(legend.text.align = 0,
 legend.box.just = "left")
netp1A

net1M <- make_network(pseqtestM, type="samples", distance="bray", max.dist=0.7, 
keep.isolates=TRUE)
netp1M <- plot_network(net1M, pseqtestM, type="samples", color="Date", 
#shape="Location",
 point_size=3, label=NA)+
 theme(legend.text.align = 0,
 legend.box.just = "left")
netp1M
```

## Alpha Diversity

```{r}
setwd("C:/Users/saysh/OneDrive/Desktop/Kuehn_Final_Analyses/Typha/Typha_Fungi/Typha_Fungi_Rerun_March2026")
alphap1 <- ggplot(alpha, aes(x=Date, y=Observed)) +
 geom_boxplot(aes(fill=Location)) +
 scale_fill_grey(start=0.9, end=0.35) +
 ylab("Species Observed")
alphap1

alphap2 <- ggplot(alpha, aes(x=Date, y=Chao1)) +
 geom_boxplot(aes(fill=Location)) +
 scale_fill_grey(start=0.9, end=0.35) +
 ylab("Chao1")
alphap2

alphap3 <- ggplot(alpha, aes(x=Date, y=Shannon)) +
 geom_boxplot(aes(fill=Location)) +
 scale_fill_grey(start=0.9, end=0.35) +
 ylab("Shannon")
alphap3

alphap4 <- ggplot(alpha, aes(x=Date, y=InvSimpson)) +
 geom_boxplot(aes(fill=Location)) +
 scale_fill_grey(start=0.9, end=0.35) +
 ylab("Inverse Simpson")
alphap4

alphaplot <- ggarrange(alphap1,alphap2,alphap3,alphap4, labels="AUTO", 
ncol=2, nrow=2, common.legend = TRUE, legend="bottom")
alphaplot

```


###BARPLOT
Check who is abundant across samples…
```{r}
setwd("C:/Users/saysh/OneDrive/Desktop/Kuehn_Final_Analyses/Typha/Typha_Fungi/Typha_Fungi_Rerun_March2026")
tax_avgs <- rowSums(major_taxa_proportions_tab)/36
#tax_avgs <- rowSums(taxmerge)/36
tax_avgs
```
```{r}
setwd("C:/Users/saysh/OneDrive/Desktop/Kuehn_Final_Analyses/Typha/Typha_Fungi/Typha_Fungi_Rerun_March2026")
#Phylum
#Select taxa that make up >90% of the total sequences
rownames(major_taxa_proportions_tab)
phylatrim <- data.frame(major_taxa_proportions_tab[c(
  "Unclassified_Phylum", "Arthoniomycetes","Dothideomycetes",
  "Eurotiomycetes", "Sordariomycetes","Unclassified_Ascomycota",
  "Agaricomycetes", "Cystobasidiomycetes", "Tremellomycetes"), ])

#Create an "other" category for the phyla not retained
filtered_proportions <- colSums(major_taxa_proportions_tab) - 
  colSums(phylatrim)

#filtered_proportions <- colSums(phy_and_class) - 
#  colSums(phylatrim)
phylatrim <- rbind(phylatrim, "Other"=filtered_proportions)
phylatrim

#Add taxa names as a column
phylatrim$Major_Taxa <- row.names(phylatrim)
#transform into long format
phylalong <- gather(phylatrim, Sample, Proportion, -Major_Taxa)
#add metadata
names(metatest)
## organize data frame
phylamet<-data.frame("Sample"=row.names(metatest),
                     "Date"=metatest$Date,
                     "Location"=metatest$Location,
                     stringsAsFactors=F)
#merge metadata with major taxa data
phylalong <- merge(phylalong, phylamet)
#Summarize by depth and hydration
phyla_summary <- 
  phylalong %>% # the names of the new data frame and the data frame to be summarised
  group_by(Date, Location, Major_Taxa) %>%   # the grouping variable
  summarise(mean_prop = mean(Proportion))  # calculates the mean of each group
## `summarise()` has grouped output by 'Depth', 'Hydration'. You can override using the `.groups` argument.
phyla_summary

write.csv(phyla_summary, "Typha_Class_Relative_Abundance_062626.csv")

phyla_summary$Date <- factor(phyla_summary$Date, levels=c("November","March","July")) #reorder variables
phyla_summary$Location <- factor(phyla_summary$Location, levels=c("Michigan","Alabama")) #reorder variables
phyla_summary$Major_Taxa <- factor(phyla_summary$Major_Taxa, levels=rev(c("Other", "Tremellomycetes", "Cystobasidiomycetes", "Agaricomycetes", "Unclassified_Ascomycota", "Sordariomycetes", "Eurotiomycetes", "Dothideomycetes", "Arthoniomycetes", "Unclassified_Phylum")))
```


```{r}
setwd("C:/Users/saysh/OneDrive/Desktop/Kuehn_Final_Analyses/Typha/Typha_Fungi/Typha_Fungi_Rerun_March2026")
#color palette
pal14 <- rev(cols25(n=14)) #set n= to the number of taxa you are plotting
#pal <- rev("colorspace::rainbow_hcl")
#make stacked bar plot
phylum_bar <- ggplot(phyla_summary, aes(x = Date, y = mean_prop, fill = Major_Taxa))+
  geom_bar(stat = "identity", col=I("black")) +
  scale_fill_manual(values=pal14)+
#  scale_color_paletteer_d("nationalparkcolors::Acadia") +
  guides(fill=guide_legend(ncol=1))+
  facet_wrap(~Location, labeller = label_both, nrow=1, scales="free_x") +
  labs(x="Date", y="Relative Abundance (%)",
       fill="Major Taxa")+
  theme(axis.text.x = element_text(angle=60, hjust=1), legend.position = "right")+
  scale_x_discrete(labels=c("November"="Nov", "March"="Mar", "July"="Jul"))
phylum_bar
 
c(180,100) *
  0.0394 * # convert mm to inch
  600 # convert to pixels
## [1] 4255.2 2364.0
plotout <- "PhylumBarplot_062626.tiff"
agg_tiff(filename=plotout, width=4255, height=2364, units="px",
         pointsize=10, res=600, compression="lzw", scaling=0.5)
phylum_bar
invisible(dev.off())
```

